A comparison of single-cell trajectory inference methods

### 2018-11-12 manuscript/paper.Rmd

`www.github.com/dynverse/dynbenchmark/tree/`r git2r::revparse_single(".","HEAD")$sha``
```{r}
n_tools <- nrow(tools)
n_tools_evaluated <- nrow(filter(tools, evaluated))
n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% "tool" | id == "paga_tree"))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
n_datasets_real <- nrow(list_datasets("real"))
n_datasets_synthetic <- nrow(list_datasets("synthetic"))
```
# Abstract

@@

In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results highlight the complementarity of existing tools, and that there is no generally best performing method. Instead, the choice of method →mostly depends on the dataset characteristics and the trajectory topology present. To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments, and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

```{r}
add_fig(
  fig_path = raw_file("overview_evaluation_v6.svg", ""),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity. d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory: the topology, the branch assignment, the cell positions and the important features.←",
   width = 13,
  height = 8, 
)
```

→
# Results
## Trajectory inference methods
<!-- Method characterisation -->
In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`). When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`). Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[a][b]

@@

<!-- Overall evaluation results -->
We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g. genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).

Unfortunately, few methods seem to have been given consideration to the scalability or the usability of the tool. While less important than the accuracy, with increasing dataset sizes the scalability of the tool is becoming increasingly important, and a high usability score is important to ensure correct usage of the tool by a user.

Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score. As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[c][d]

@@

```{r}
add_sfig(
  fig_path = result_file("results_suppfig.pdf", experiment_id = "08-summary"),
  ref_id = "results", 
  caption_main = glue::glue(""), 
  caption_text = glue::glue(""),
  width = 30,
  height = 18,
  integrate = FALSE
)
```
## Accuracy
<!-- Accuracy methods -->
We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose four metrics each assessing a different aspect of a trajectory (`r ref("fig", "ti_evaluation_overview", "d")`): the topology (Hamming-Ipsen-Mikhailov, `r label_metric("him")`), the quality of clustering of the assignment of cells to branches (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).  Some real datasets were classified as "gold standard" if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing [@tian_scrnaseqmixologybetter_2018]. All other real datasets were classified as "silver standard". For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation[@schaffter_genenetweaversilicobenchmark_2011]. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties[@zappia_splattersimulationsinglecell_2017].

<!-- Benchmark overview -->
We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`). Even methods which can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "b")`). The overall score between the different dataset sources was moderately to highly correlated (0.5 - 0.9) with the scores on real datasets with a gold standard (`r ref("sfig", "benchmark_interpretation", "b")`), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data. On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "b")`).

@@

<!-- Complementarity between methods -->
The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods. A top model in this case was defined as a model with an overall score at least 95% as the best model. On all datasets, using one method would result in getting a top model in about 40% of the time. This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`). This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves, and methods with a broad trajectory type range such as PAGA. We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to obtain at least one top model for a given dataset. Altogether, this shows that there is considerable complementarity between the different methods, and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori. Moreover, it also opens up the possibilities for new ensemble methods which utilize this complementarity.

@@

## Scalability
While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million [@svensson_exponentialscalingsinglecell_2018]. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features [@cao_jointprofilingchromatin_2018]. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We then modelled the running time using a generalized additive model[@wood_thinplateregression_2003] (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`). As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[e][f]

Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions stand out: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree). Despite their ability to detect all trajectory types, these two methods are still able to accurately detect a trajectory with 100.000 cells and 10.000 features in less than 10 minutes (`r ref("fig", "results_detailed", "b")`). Methods with a low running time typically have two defining aspects: they have a sublinear time complexity with respect to the features and/or cells, and adding new cells or features leads to a relatively low increase in time (`r ref("sfig", "scaling", "b")`). We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).

```{r echo=FALSE, results="asis"}
add_sfig(
  fig_path = result_file("scaling.pdf", experiment_id = "05-scaling"),
  ref_id = "scaling", 
  caption_main = "Scalability of trajectory inference methods", 
  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right). (b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We used to GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve. We also used the predicted timings to calculate the average increase per new 100 cells or features. Also shown is the average maximal memory usage.",
  width = 12,
  height = 15
)
```
## Stability
It is not only important that a method is able to infer an accurate model, but also that it produces more-or-less the same model when the input data is perturbed slightly. In order to test the stability of each method, we executed each method on 10 different bootstraps of the datasets (95% cells, 95% features), and calculated the average similarity between each pair of models (`r ref("fig", "results_detailed", "d")`).

@@

We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

```{r}
add_sfig(
  fig_path = result_file("qc_overview.rds", experiment_id = "03-methods/02-tool_qc"),
  ref_id = "qc_overview", 
  caption_main = "Usability of trajectory inference methods.", 
  caption_text = glue::glue("Shown is the score given for each method on every item from our usability score sheet ({ref('stable', 'qc_scoresheet')}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average QC score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had an average score lower than 0.5."),
   width = 15,
   height = 18
)
```

@@

________________
# Online Methods
## Data and code availability
→ 
The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[g])) [@cannoodt_goldstandardsinglecell_].

@@

Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[h][i][j][k]
←
## Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task. During the execution of a method on a dataset, if the time limit (>1h) or memory limit (16GB) was exceeded, or an error was produced, a zero score was returned for that execution.
→
## Complementarity
To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score is equal or higher than 95% of the best overall score for that particular dataset. We then calculated for every method the weighted percentage of datasets which fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method. We iteratively added new methods until all methods were selected. For this analysis, we did not include any methods which require prior information, and only included methods which can detect at least one of the trajectory types present in the datasets.
## Scalability
To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section. We up- and downscaled these datasets between 10 to 100.000 cells and 10 to 100.000 features, while never going higher than 1.000.000 values in total. To generate new cells or features, we first generated a 10-nearest neighbour graph of both the cells and features from the expression space. For every new cell or feature, we used a linear combination of one to three existing cells or features, where each cell or features was given a weight sampled from $\textit{U}(0,1)$.

@@

## Stability[l][m]
We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking. We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features, such that each method should produce slightly different results. 
←
## Usability

@@

[a]Try to rewrite a little bi
[b]somehing like his?
[c]TODO: rework enumeration into a paragraph
[d]something like this?
[e](Not shown), or include in Supp Fig 4?
[f]Yeah we don't include this yet, should be in sup fig 4
[g]Update when final
[h]This might give the impression of being a bit "overengineered".  Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
[i]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
[j]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
[k]I adapted the text a little bit to better explain the 'why'
[l]Todo once this experiment is finished
[m]Good enough?

>>>>

Commit: `r print_url(paste0("www.github.com/dynverse/dynbenchmark/tree/", git2r::revparse_single(".","HEAD")$sha))`
```{r}
n_tools <- nrow(tools)
n_tools_evaluated <- nrow(filter(tools, evaluated))
n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% "tool" | id == "paga_tree"))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
n_datasets_real <- nrow(list_datasets("real"))
n_datasets_synthetic <- nrow(list_datasets("synthetic"))
```
# Abstract

@@

In this study, we evaluated a total of →`r n_methods_evaluated`← m[a]ethods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results highlight the complementarity of existing tools, and that there is no generally best performing method. Instead, the choice of method →mostly depends on the dataset characteristics and the trajectory topology present. To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments, and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

```{r}
add_fig(
  fig_path = raw_file("overview_evaluation_v6.svg", ""),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation[b].", 
  caption_text = glue::glue("a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity. d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory. The Hamming-Ipsen-Mikhailov (HIM) score will assess the similarity between the two topologies, taking into account differences in edge lengths and degree distributions. The {label_metric('F1_branches')} assesses the similarity of the assignment of cells onto branches. The {label_metric('correlation')} quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances. Finally, {label_metric('featureimp_wcor')} quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.←"),
   width = 14,
  height = 8.5, 
)
```

→
# Results
## Trajectory inference methods
<!-- Method characterisation -->
In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated`[c] TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`). When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`). Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[d][e][f]

@@

<!-- Overall evaluation results -->
We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real`[g] real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g. genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).

Unfortunately, only few methods seem to have been given consideration to the scalability or the usability of the tool. While less important than the accuracy, with increasing dataset sizes the scalability of the tool is becoming increasingly important, and a high usability score is important to ensure correct usage of the tool by a user.

Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score. As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[h][i]

@@

```{r}
add_sfig(
  fig_path = result_file("results_suppfig.pdf", experiment_id = "08-summary"),
  ref_id = "results", 
  caption_main = glue::glue(""), 
  caption_text = glue::glue(""),
  width = 30,
  height = 18,
  integrate = FALSE
)
```
## Accuracy
<!-- Accuracy methods -->
We defined several metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose four metrics each assessing a different aspect of a trajectory (`r ref("fig", "ti_evaluation_overview", "d")`): the topology (Hamming-Ipsen-Mikhailov, `r label_metric("him")`), the quality of the assignment of cells to branches (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the accuracy of the differentially expressed features along the trajectory (`r label_metric("featureimp_wcor")`). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).  Some real datasets were classified as "gold standard" if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing [@tian_scrnaseqmixologybetter_2018]. All other real datasets were classified as "silver standard". For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation[@schaffter_genenetweaversilicobenchmark_2011]. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties[@zappia_splattersimulationsinglecell_2017].

<!-- Benchmark overview -->
We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`). Even methods which can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "b")`). The overall score between the different dataset sources was moderately to highly correlated (0.5 - 0.9) with the scores on real datasets containing a gold standard (`r ref("sfig", "benchmark_interpretation", "b")`), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data. On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "b")`).

@@

<!-- Complementarity between methods -->
The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods. A top model in this case was defined as a model with an overall score of at least 95% as the best model. On all datasets, using one method would result in getting a top model in about 40% of the time. This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`). This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves[j][k], and methods with a broad trajectory type range such as PAGA. We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to obtain at least one top model for a given dataset. Altogether, this shows that there is considerable complementarity between the different methods, and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori. Moreover, it also opens up the possibilities for new ensemble methods which utilize this complementarity.

@@

## Scalability
While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million [@svensson_exponentialscalingsinglecell_2018]. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features [@cao_jointprofilingchromatin_2018]. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modelled the running time and memory usage using a Shape Constrained Additive Model[@pya_shapeconstrainedadditive_2015] (`r ref("sfig", "scaling", "a")`). We assessed whether the time / memory complexity was sublinear, linear ,quadratic, or superquadratic with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`). As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[l][m][n]

Overall, methods which detected more complex trajectory types also had longer running times, except for PAGA and PAGA Tree (`r ref("fig", "results_detailed", "c")`). We looked at the running times of some common use cases, such as datasets with a lot of cells (100 thousand to 1 million) but only a limited number of features (100 - 1000), which is usually the case for droplet based datasets, and datasets with only a limited number of cells but a lot of features, relevant for flow and mass cytometry datasets[o]. Only a handful of methods had an acceptable running time on these datasets, with PAGA tree, GrandPrix and several linear methods able to finish in under 10 minutes on a dataset with one hundred thousand cells and one thousand features. Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (`r ref("sfig", "scaling", "b")`). We found that more than half of all methods had a linear or superlinear complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).

```{r echo=FALSE, results="asis"}
add_sfig(
  fig_path = result_file("scaling.pdf", experiment_id = "05-scaling"),
  ref_id = "scaling", 
  caption_main = "Scalability of trajectory inference methods", 
  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right). (b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We used the GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve. We also used the predicted timings to calculate the average increase per new 100 cells or features. Also shown is the average maximal memory usage.",
  width = 12,
  height = 15
)
```
## Stability
It is not only important that a method is able to infer an accurate model, but also that it produces more-or-less the same model when the input data is perturbed slightly. In order to test the stability of each method, we executed each method on 10 different bootstraps of the datasets (95% cells, 95% features), and calculated the average similarity between each pair of models (`r ref("fig", "results_detailed", "d")`).

@@

We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

```{r}
add_sfig(
  fig_path = result_file("qc_overview.rds", experiment_id = "03-methods/02-tool_qc"),
  ref_id = "qc_overview", 
  caption_main = "Usability of trajectory inference methods.", 
  caption_text = glue::glue("Shown is the score given for each method on every item from our usability score sheet ({ref('stable', 'qc_scoresheet')}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average usability score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had an average score lower than 0.5."),
   width = 15,
   height = 18
)
```

@@

________________
# Online Methods
## Data and code availability
→ 
The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[p])) [@cannoodt_goldstandardsinglecell_].

@@

Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[q][r][s][t]
←
## Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task. During the execution of a method on a dataset, if the time limit (>1h) or memory limit (16GB) was exceeded, or an error was produced, a zero score was returned for that execution.
→
## Complementarity
To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score is equal or higher than 95% of the best overall score for that particular dataset. We then calculated for every method the weighted percentage of datasets which fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method. We iteratively added new methods until all methods were selected. For this analysis, we did not include any methods which require prior information, and only included methods which can detect at least one of the trajectory types present in the datasets.
## Scalability
To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section. We up- and downscaled these datasets between 10 to 100.000 cells and 10 to 100.000 features, while never going higher than 1.000.000 values in total. To generate new cells or features, we first generated a 10-nearest neighbour graph of both the cells and features from the expression space. For every new cell or feature, we used a linear combination of one to three existing cells or features, where each cell or features was given a weight sampled from $\textit{U}(0,1)$.

@@

## Stability[u][v]
We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking. We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features, such that each method should produce slightly different results. 
←
## Usability

@@

[a]Check that this is now without the control methods
[b]Mentions "55" methods instead of the "45" mentioned in the introduction. Also, the things written in the figure b are very small.
[c]check consistency with Figure 1
[d]Try to rewrite a little bi
[e]somehing like his?
[f]Hehe ;). Something like this?
[g]Does not fit with the number in Figure 1
[h]TODO: rework enumeration into a paragraph
[i]something like this?
[j]Isn't this a control method ?
[k]yes, should be removed!
[l](Not shown), or include in Supp Fig 4?
[m]Yeah we don't include this yet, should be in sup fig 4
[n]TODO: Will be added as sup fig
[o]I would not mention flow/mass cytoemtry here
[p]Update when final
[q]This might give the impression of being a bit "overengineered".  Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
[r]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
[s]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
[t]I adapted the text a little bit to better explain the 'why'
[u]Todo once this experiment is finished
[v]Good enough?
